---
title: "Extra Demonstration: Datatypes and I/O"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# The below commands toggle scrolling in outputs
options(width = 60)
local({
  hook_output <- knitr::knit_hooks$get('output')
  knitr::knit_hooks$set(output = function(x, options) {
    if (!is.null(options$max.height)) options$attr.output <- c(
      options$attr.output,
      sprintf('style="max-height: %s;"', options$max.height)
    )
    hook_output(x, options)
  })
})
```

In this demonstration we will practice with data formats encountered in medical records.

#### Scope

This example includes extra material that we did not cover during this iteration of the course. We will demonstrate the following...

1. Loading data into data frames
2. Formatting time stamps
3. Performing duration arithmetic
4. Writing outputs to a CSV file

Since this material was not covered in the course, there are no fill-in-the-blanks associated with this notebook.

#### Data Description

We will be working with synthetic patient records generated with the Synthea$^{\text{TM}}$ software package made available by The MITRE Corporation. The synthetic data models 1000 patient records and is stored in 16 CSV tables. The data source is cited below:

>Jason Walonoski, Mark Kramer, Joseph Nichols, Andre Quina, Chris Moesel, Dylan Hall, Carlton Duffett, Kudakwashe Dube, Thomas Gallagher, Scott McLachlan, Synthea: An approach, method, and software mechanism for generating synthetic patients and the synthetic electronic health care record, Journal of the American Medical Informatics Association, Volume 25, Issue 3, March 2018, Pages 230–238, [https://doi.org/10.1093/jamia/ocx079](https://doi.org/10.1093/jamia/ocx079)

The data may be downloaded by clicking [this](https://synthetichealth.github.io/synthea-sample-data/downloads/synthea_sample_data_csv_apr2020.zip) hyperlink. **Please ensure that all tables are stored subdirectory called "data" placed immediately inside the directory containing this notebook.**

#### Import Modules

```{r}
library(tidyverse)
```

#### Specify and Load Data

In this demonstration, we will be working with just one data table containing patient encounters. This data file is located at `data/encounters.csv`, relative to the directory containing this notebook. Below, we load this table from the CSV file into a data frame.

```{r, max.height='200px'}
# Specify the path to the file containing patient encounters
encounters_path <- "data/encounters.csv"

# Load the encounters data into a data frame
encounters <- read.csv(encounters_path, header = TRUE, stringsAsFactors = FALSE)

# Display the table
print(encounters)
```

#### Format Date/Time information

The `encounters` table includes `START` and `STOP` timestamps denoting the start and end times of each encounter. We format these columns into date/time format using the `POSIXct` package.

```{r}
# Convert `START` and `STOP` columns to date
encounters = encounters %>%
  mutate(START = as.POSIXct(START, format = "%Y-%m-%dT%H:%M:%SZ")) %>%
  mutate(STOP = as.POSIXct(STOP, format = "%Y-%m-%dT%H:%M:%SZ"))
```

#### Task 3: Identify the Duration of Each Encounter

For each encounter, we take the difference between the values in the `STOP` and `START` columns and express the result in units of hours. We store the results in a new column of the `encounters` table called `duration (hr)`.

```{r, max.height='200px'}
# We take the difference between the `STOP` and `START` columns and store the result in a `duration column
# We then convert the duration to units of hours
encounters = encounters %>%
  mutate(duration = difftime(STOP, START, units="secs")) %>%
  mutate(duration_hr = as.numeric(duration) / 3600)
print(encounters)
```

#### Plot a Histogram of Encounter Durations

We visualize the distribution of encounter durations. Notice, there will be some outliers. To address outliers, we exclude durations greater than four hours from your histogram.

```{r}
# Isolate the `duration_hr` column in its own vector and remove values greater than 4
durations_hr_vec <- encounters %>%
  select(duration_hr) %>%
  filter(duration_hr <= 4) %>%
  pull()

# Plot a histogram of the encounter durations
bin_edges <- seq(0, 4, length.out = 41)
ggplot(data = data.frame(durations_hr_vec), aes(x = durations_hr_vec)) +
  geom_histogram(binwidth = 0.1, color = "black", fill = "steelblue", breaks = bin_edges) +
  xlab("Encounter Duration (hrs)") +
  ylab("Frequency")
```

It appears that the durations are often recorded in 15 minute intervals, with a minimum encounter duration of 15 minutes.

#### Isolate the Encounters for a Single Patient

We isolate the encounters associated with patient `034e9e3b-2def-4559-bb2a-7850888ae060`. To do so, we filter the rows in the `encounters` table to those associated with our patient of interest.

```{r}
# Specify a patient of interest
patient <- "034e9e3b-2def-4559-bb2a-7850888ae060"

# Filter the `encounters` table to rows with `034e9e3b-2def-4559-bb2a-7850888ae060` in the `PATIENT` column
patient_encounters <- encounters %>%
  filter(PATIENT == patient)

# Sort the patient encounters by the `START` times
patient_encounters <- patient_encounters %>%
  arrange(START)
```

```{r}
# How many encounters did the patient of interest have
n_encounters <- nrow(patient_encounters)

# Print the number of encounters associated with the patient
cat(paste0("The patient had ", n_encounters, " encounters."))
```

#### Identify Medical Costs Following an Encounter

Medical follow-up can be costly! We will work through an analysis of the six-month follow-up costs after each procedure. Given an encounter, we will implement a function to do the following:

1. Identify the date six months after the `STOP` of the encounter
2. Identify all subsequent encounters for the same patient during the six-month follow-up period
3. Calculate the total medical cost associated with the follow-up period

```{r}
# Create a function to identify the cost of follow-up encounters during a six-month period following an encounter
cost_of_follow_ups <- function(encounter) {
  
  # Identify the STOP time of the encounter
  encounter_stop_time <- as.POSIXlt(encounter$STOP)
  
  # Identify the time six months (~182 days) following the encounter
  six_months_later <- encounter_stop_time + as.difftime(182, units = "days")
  
  # Identify the patient associated with the encounter
  patient_id <- encounter$PATIENT
  
  # Filter the `encounters` table to rows matching the patient
  patient_specific_encounters <- encounters %>%
    filter(PATIENT == patient_id)
  
  # Filter the `patient_specific_encounters` to encounters with START times in the six-month period
  patient_specific_encounters <- patient_specific_encounters %>%
    filter(START > encounter_stop_time, START < six_months_later)
  
  # Evaluate the total cost of the follow-ups
  tot_cost <- sum(patient_specific_encounters$TOTAL_CLAIM_COST)
  return(tot_cost)
}
```

```{r}
# Test the `cost_of_follow_ups` on specific encounters
n_eval <- 10
for (i in 1:n_eval) {
  encounter <- encounters[sample(nrow(encounters), 1), ]
  
  patient_id <- encounter$PATIENT
  code <- encounter$CODE
  follow_up_costs <- cost_of_follow_ups(encounter)
  
  cat("Patient:", patient_id, "\n")
  cat("Encounter Code:", code, "\n")
  cat("Follow-up Costs:", follow_up_costs, "\n\n")
}
```

#### Task 8: Summarize the Costs of Patient Encounters and Export to CSV

We generate descriptive statistics summarizing the costs associated with encounters, and export those summary statistics to a CSV file.

```{r}
# List the columns associated with some cost
cost_cols <- c("BASE_ENCOUNTER_COST", "TOTAL_CLAIM_COST", "PAYER_COVERAGE")

# Isolate the costs associated with patient encounters
encounter_costs <- encounters[cost_cols]

# Generate a table summarizing each of the costs columns with descriptive statistics
encounter_cost_summaries <- summary(encounter_costs)
print(encounter_cost_summaries)
```

```{r}
# Specify a path at which to save the encounter summaries
save_path <- "data/encounter_cost_summaries.csv"

# Save the `encounter_cost_summaries` to a CSV file
write.csv(encounter_cost_summaries, file = save_path, row.names = FALSE)

# Read the CSV file that you just saved
encounter_cost_summaries_reloaded <- read.csv(save_path, header = TRUE)

# Display the reloaded table summarizing the costs of encounters
print(encounter_cost_summaries_reloaded)
```

#### Summary

In this notebook, we worked with date/time data to summarize patient encounters reported in medical records. After loading and formatting the data, we calculated the duration of encounters and plotted the distribution of encounter durations. Given a specific encounter, we implemented a function to identify follow-up encounters and calculate the total cost of those follow-ups. Finally, we generated summary statistics that describe the cost of encounters, and we exported those summary statistics to a CSV file.
