---
title: "Extra Demonstration: Datatypes and I/O"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# The below commands toggle scrolling in outputs
options(width = 60)
local({
  hook_output <- knitr::knit_hooks$get('output')
  knitr::knit_hooks$set(output = function(x, options) {
    if (!is.null(options$max.height)) options$attr.output <- c(
      options$attr.output,
      sprintf('style="max-height: %s;"', options$max.height)
    )
    hook_output(x, options)
  })
})
```

In this demonstration we will practice with data formats encountered in health records.

#### Scope

This example includes extra material that we did not cover during this iteration of the course. We will demonstrate the following...

1. Loading data into data frames
2. Formatting time stamps
3. Performing duration arithmetic
4. Writing outputs to a CSV file

Since this material was not covered in the course, there are no fill-in-the-blanks associated with this notebook.

#### Data Description

We will be working with data from NHANES, the National Health and Nutrition Examination Survey. NHANES has been run by the CDC since the 1960s. Every two years it interviews a nationally representative sample of about 10,000 Americans, brings them into a mobile examination center for a physical exam, and draws blood. We are using the 2017-2018 cycle, which examined 9,254 people.

The public NHANES files are documented here:

>Centers for Disease Control and Prevention (CDC). National Center for Health Statistics (NCHS). National Health and Nutrition Examination Survey Data, 2017-2018. Hyattsville, MD: U.S. Department of Health and Human Services, Centers for Disease Control and Prevention. [https://wwwn.cdc.gov/nchs/nhanes/continuousnhanes/default.aspx?BeginYear=2017](https://wwwn.cdc.gov/nchs/nhanes/continuousnhanes/default.aspx?BeginYear=2017)

The tables we use here were built from those files by `data/build_nhanes_data.R` in this repository. **The tables are already stored in a subdirectory called "data" placed immediately inside the directory containing this notebook.** Note that NHANES does not release exam dates or session times, so the `exam_start` and `exam_end` columns are simulated (each participant's exam falls inside their real reported examination period). Every measurement, lab value, and drug name is real.

#### Import Modules

```{r}
library(tidyverse)
```

#### Specify and Load Data

In this demonstration, we will be working with just one data table containing exam components. This data file is located at `data/exams.csv`, relative to the directory containing this notebook. Below, we load this table from the CSV file into a data frame.

```{r, max.height='200px'}
# Specify the path to the file containing exam components
exams_path <- "data/exams.csv"

# Load the exams data into a data frame
exams <- read.csv(exams_path, header = TRUE, stringsAsFactors = FALSE)

# Display the table
print(exams)
```

#### Format Date/Time information

The `exams` table includes `exam_start` and `exam_end` timestamps denoting the start and end times of each exam component. We format these columns into date/time format using the `POSIXct` package.

```{r}
# Convert `exam_start` and `exam_end` columns to date
exams = exams %>%
  mutate(exam_start = as.POSIXct(exam_start, format = "%Y-%m-%dT%H:%M:%SZ")) %>%
  mutate(exam_end = as.POSIXct(exam_end, format = "%Y-%m-%dT%H:%M:%SZ"))
```

#### Task 3: Identify the Duration of Each Exam Component

For each exam component, we take the difference between the values in the `exam_end` and `exam_start` columns and express the result in units of hours. We store the results in a new column of the `exams` table called `duration_hr`.

```{r, max.height='200px'}
# We take the difference between the `exam_end` and `exam_start` columns and store the result in a `duration` column
# We then convert the duration to units of hours
exams = exams %>%
  mutate(duration = difftime(exam_end, exam_start, units="secs")) %>%
  mutate(duration_hr = as.numeric(duration) / 3600)
print(exams)
```

#### Plot a Histogram of Component Durations

We visualize the distribution of component durations. Notice, there will be some outliers. To address outliers, we exclude durations greater than four hours from your histogram.

```{r}
# Isolate the `duration_hr` column in its own vector and remove values greater than 4
durations_hr_vec <- exams %>%
  select(duration_hr) %>%
  filter(duration_hr <= 4) %>%
  pull()

# Plot a histogram of the component durations
bin_edges <- seq(0, 4, length.out = 41)
ggplot(data = data.frame(durations_hr_vec), aes(x = durations_hr_vec)) +
  geom_histogram(binwidth = 0.1, color = "black", fill = "steelblue", breaks = bin_edges) +
  xlab("Component Duration (hrs)") +
  ylab("Frequency")
```

It appears that the durations are recorded in 5 minute intervals, with a minimum component duration of 20 minutes.

#### Isolate the Exam Components for a Single Participant

We isolate the exam components associated with participant `NHANES-100000`. To do so, we filter the rows in the `exams` table to those associated with our participant of interest.

```{r}
# Specify a participant of interest
participant <- "NHANES-100000"

# Filter the `exams` table to rows with `NHANES-100000` in the `participant_id` column
participant_exams <- exams %>%
  filter(participant_id == participant)

# Sort the exam components by the `exam_start` times
participant_exams <- participant_exams %>%
  arrange(exam_start)
```

```{r}
# How many exam components did the participant of interest have
n_exams <- nrow(participant_exams)

# Print the number of exam components associated with the participant
cat(paste0("The participant had ", n_exams, " exam components."))
```

#### Identify Examination-Center Workload Following a Component

A mobile examination center has a finite amount of time on the road, so knowing how heavily each center is booked matters for planning. We will work through an analysis of the week of work that follows any given component. Given a component, we will implement a function to do the following:

1. Identify the date one week after the `exam_end` of the component
2. Identify all subsequent components run at the same center during that week
3. Calculate the total examination time booked during that week

```{r}
# Create a function to identify the examination time booked in the week following a component
hours_of_follow_up_work <- function(component) {

  # Identify the exam_end time of the component
  component_stop_time <- as.POSIXlt(component$exam_end)

  # Identify the time one week (7 days) following the component
  one_week_later <- component_stop_time + as.difftime(7, units = "days")

  # Identify the examination center associated with the component
  center <- component$center_id

  # Filter the `exams` table to rows matching the center
  center_specific_exams <- exams %>%
    filter(center_id == center)

  # Filter the `center_specific_exams` to components with exam_start times in the one-week period
  center_specific_exams <- center_specific_exams %>%
    filter(exam_start > component_stop_time, exam_start < one_week_later)

  # Evaluate the total examination time booked in that week
  tot_hours <- sum(center_specific_exams$duration_hr)
  return(tot_hours)
}
```

```{r}
# Test the `hours_of_follow_up_work` on specific components
n_eval <- 10
for (i in 1:n_eval) {
  component <- exams[sample(nrow(exams), 1), ]

  participant_id <- component$participant_id
  which_component <- component$component
  follow_up_hours <- hours_of_follow_up_work(component)

  cat("Participant:", participant_id, "\n")
  cat("Component:", which_component, "\n")
  cat("Follow-up work (hrs):", follow_up_hours, "\n\n")
}
```

#### Task 8: Summarize the Exam Components and Export to CSV

We generate descriptive statistics summarizing the exam components, and export those summary statistics to a CSV file.

```{r}
# List the numeric columns we want to describe
numeric_cols <- c("duration_hr", "status_code")

# Isolate those columns
exam_numbers <- exams[numeric_cols]

# Generate a table summarizing each of those columns with descriptive statistics
exam_summaries <- summary(exam_numbers)
print(exam_summaries)
```

```{r}
# Specify a path at which to save the exam summaries
save_path <- "data/exam_summaries.csv"

# Save the `exam_summaries` to a CSV file
write.csv(exam_summaries, file = save_path, row.names = FALSE)

# Read the CSV file that you just saved
exam_summaries_reloaded <- read.csv(save_path, header = TRUE)

# Display the reloaded table summarizing the exam components
print(exam_summaries_reloaded)
```

#### Summary

In this notebook, we worked with date/time data to summarize exam components reported in health records. After loading and formatting the data, we calculated the duration of components and plotted the distribution of those durations. Given a specific component, we implemented a function to identify subsequent components at the same examination center and calculate the total examination time booked in the following week. Finally, we generated summary statistics that describe the components, and we exported those summary statistics to a CSV file.
