---
title: "Practice 2: R Fundamentals"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# The below commands toggle scrolling in outputs.
# Please do not adjust this code!
options(width = 60)
local({
  hook_output <- knitr::knit_hooks$get('output')
  knitr::knit_hooks$set(output = function(x, options) {
    if (!is.null(options$max.height)) options$attr.output <- c(
      options$attr.output,
      sprintf('style="max-height: %s;"', options$max.height)
    )
    hook_output(x, options)
  })
})
```

This notebook provides the opportunity to practice loading and manipulating data from yesterday's healthcare data set.

#### Instructions

This notebook includes a framework for inspecting data in the example healthcare data set. Skeleton code for the analysis is provided for you. Your job is to fill in missing code denoted by ellipses (`...`) in the notebook. Each ellipse includes a commented `TODO` statement above it with instructions. Please fill in the missing code in the notebook.

#### Scope

This notebook includes material covered in Lecture 2: R Fundamentals. You will have a chance to...

1. Assign variables
2. Load and inspect data frames

#### Data Description

We will be working with synthetic patient records generated with the Synthea$^{\text{TM}}$ software package made available by The MITRE Corporation. The synthetic data models 1000 patient records and is stored in 16 CSV tables. The data source is cited below:

>Jason Walonoski, Mark Kramer, Joseph Nichols, Andre Quina, Chris Moesel, Dylan Hall, Carlton Duffett, Kudakwashe Dube, Thomas Gallagher, Scott McLachlan, Synthea: An approach, method, and software mechanism for generating synthetic patients and the synthetic electronic health care record, Journal of the American Medical Informatics Association, Volume 25, Issue 3, March 2018, Pages 230–238, [https://doi.org/10.1093/jamia/ocx079](https://doi.org/10.1093/jamia/ocx079)

The data may be downloaded by clicking [this](https://synthetichealth.github.io/synthea-sample-data/downloads/synthea_sample_data_csv_apr2020.zip) hyperlink. **Please ensure that all tables are stored subdirectory called "data" placed immediately inside the directory containing this notebook.**

#### Import Modules

```{r}
library(ggplot2)
```

#### Specify and Load Data

During the first demonstration, we loaded patient data into a data frame. Please load the table below using the `read.csv` function. If you are stumped, please look back at the first demonstration or review the documentation [here](https://www.rdocumentation.org/packages/qtl2/versions/0.32/topics/read_csv).

```{r}
# TODO: Assign the value "data/patients.csv" to the variable `path`
path <- ...

# TODO: Load the table using read.csv
df_patient <- ...
```

#### Load Column Names

A data frame includes columns of fixed data types that are labeled with headers. You can identify the headers in a data frame using the `colnames` function. After reviewing the documentation for this function [here](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/row%2Bcolnames), please use the `colnames` function to extract the column names in the `df_patient` data frame.

```{r}
# TODO: Extract the column names of the `df_patient` table using the `colnames` function
col_names <- ...

# Print the column names
print(col_names)
```

#### Identify Unique Patients

The built-in `unique` function is used to identify unique values in a data set. If the function is applied to a data frame, then unique rows will be returned. If the function is applied to a vector, then unique elements will be returned. Please review the documentation for the `unique` function [here](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/unique), then use the function to identify the unique values in the `Id` column of `df_patient`.

```{r, max.height='200px'}
# TODO: Identify the unique values in the `Id` column of `df_patient`
# Store the unique patient IDs in a vector
unique_patients <- ...

# Print the unique patients
print(unique_patients)
```

#### Count Unique Patients

The built-in `length` function can be used to identify the number of elements in a vector. Please review the documentation for the `length` function [here](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/length), then use the function to identify the number of unique patients in `df_patient`.

```{r}
# TODO: Use the `length` function to identify the number of unique patients in `df_patient`
n_patients <- ...

# Print the number of patients (There should be 1171 unique patients)
print(n_patients)
```

#### Sort Patient Identifiers

The built-in `sort` function can be used to sort values in a vector. When the values in a vector are alphanumeric, they are sorted in alphabetical order with numbers coming first. Please review the documentation for the `sort` function [here](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/sort), then use the function to sort the unique patient identifiers.

```{r, max.height='200px'}
# TODO: Sort the patients listed in `unique_patients` by their identifiers
unique_patients_sorted <- ...

# Print the sorted patient identifiers
print(unique_patients_sorted)
```

#### Load Patient Encounters

Now, we will open a second table from our data set to inspect encounters with patients. Again using the `read.csv` function, please load the encounters data stored in `data/encounters.csv`.

```{r, max.height='200px'}
# Specify the path to the encounters data
encounters_path <- "data/encounters.csv"

# TODO: Use the `read.csv`function to load the patient encounters into a data frame
df_encounters <- ...
```

#### Define a Vector

We will use this notebook to inspect the number of patient encounters in each month. Please define a vector specifying the twelve months of the year ("January" through "December") as strings values.

```{r}
# TODO: Generate a vector of strings specifying the twelve months of the year
months <- ...
```

#### Extract Encounter Dates

Using the code provided below, we will extract the date of each encounter. We will get this information from the `START` column of the `encounters` table. We will format the encounter dates as POSIXct (date and time) information. POSIXct internally represents date and time as seconds since Junuary 1, 1970. We use the `as.POSIXct` function (with documentation [here](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/as.POSIX*)) to convert string timestamps into POSIXct format. You have no task in the subsequent code block.

```{r, max.height='200px'}
# Extract encounters for each patient
encounters <- df_encounters[, c("PATIENT", "START")]

# Reformat `START` column as `POSIXct` (datetime) information
encounters$START <- as.POSIXct(encounters$START)

# Print the encounter start times
print(encounters)
```

#### Identify the Month of Each Encounter

For each of the patient encounters, we will identify the month of that encounter. We will do so by reformatting the data in the `START` column using the `format` function (with documentation found [here](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/format)). The "%B" identifier specifies unabbreviated months (as noted [here](https://www.statmethods.net/input/dates.html). We will store the month in a new column. You have no task in the subsequent code block.

```{r, max.height='200px'}
# Extract the month for each encounter
encounters$MONTH <- format(encounters$START, "%B")

# Display the resulting table
print(encounters)
```

#### Count Month Frequencies

We will loop through each month in the `months` vector. For each month, we will do the following:

1. Filter the rows in `encounters` to those with a corresponding value in the `MONTH` column
2. Count the number of rows in the filtered table
3. Store the row count in a dictionary called `n_encounters_by_month`

This will give us the number of encounters in each month. You have no task in the subsequent code block.

```{r}
# Initialize a list for encounter counts associated with each month
n_encounters_by_month <- list()

# Loop over months
for (month in months) {
  
  # Filter rows in `encounters` matching month
  encounters_filtered <- encounters[encounters$MONTH == month, ]
  
  # Count the number of rows in `encounters_filtered`
  nrows <- nrow(encounters_filtered)
  
  # Store the row count in `n_encounters_by_month`
  n_encounters_by_month[[month]] <- nrows
  
  # Print the number of encounters associated with the month
  cat("Month:", month, "\n")
  cat("Number of encounters:", nrows, "\n\n")
}
```

#### Summary

In this notebook, we loaded patient records into a data frame. We identified and sorted the unique patient IDs in our records. We also identified and plotted the number of patient encounters by month.
