---
title: "Practice 2: R Fundamentals"
format: html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# The below commands toggle scrolling in outputs.
# Please do not adjust this code!
options(width = 60)
local({
  hook_output <- knitr::knit_hooks$get('output')
  knitr::knit_hooks$set(output = function(x, options) {
    if (!is.null(options$max.height)) options$attr.output <- c(
      options$attr.output,
      sprintf('style="max-height: %s;"', options$max.height)
    )
    hook_output(x, options)
  })
})
```

This notebook provides the opportunity to practice loading and manipulating data from yesterday's NHANES data set.

#### Instructions

This notebook includes a framework for inspecting data in the example NHANES data set. Skeleton code for the analysis is provided for you. Your job is to fill in missing code denoted by ellipses (`...`) in the notebook. Each ellipse includes a commented `TODO` statement above it with instructions. Please fill in the missing code in the notebook.

#### Scope

This notebook includes material covered in Lecture 2: R Fundamentals. You will have a chance to...

1. Assign variables
2. Load and inspect data frames

#### Data Description

We will be working with data from NHANES, the National Health and Nutrition Examination Survey. NHANES has been run by the CDC since the 1960s. Every two years it interviews a nationally representative sample of about 10,000 Americans, brings them into a mobile examination center for a physical exam, and draws blood. We are using the 2017-2018 cycle, which examined 9,254 people.

The public NHANES files are documented here:

>Centers for Disease Control and Prevention (CDC). National Center for Health Statistics (NCHS). National Health and Nutrition Examination Survey Data, 2017-2018. Hyattsville, MD: U.S. Department of Health and Human Services, Centers for Disease Control and Prevention. [https://wwwn.cdc.gov/nchs/nhanes/continuousnhanes/default.aspx?BeginYear=2017](https://wwwn.cdc.gov/nchs/nhanes/continuousnhanes/default.aspx?BeginYear=2017)

The tables we use here were built from those files by `data/build_nhanes_data.R` in this repository. **The tables are already stored in a subdirectory called "data" placed immediately inside the directory containing this notebook.** Note that NHANES does not release exam dates or dates of birth, so the `exam_start`, `exam_end`, and `birth_date` columns are simulated (consistent with each participant's real reported age and examination period). Every measurement, lab value, and drug name is real.

#### Import Modules

```{r}
library(ggplot2)
```

#### Specify and Load Data

During the first demonstration, we loaded participant data into a data frame. Please load the table below using the `read.csv` function. If you are stumped, please look back at the first demonstration or review the documentation [here](https://www.rdocumentation.org/packages/qtl2/versions/0.32/topics/read_csv).

```{r}
# TODO: Assign the value "data/participants.csv" to the variable `path`
path <- "data/participants.csv"

# TODO: Load the table using read.csv
df_participant <- read.csv(path, header = TRUE)
```

#### Load Column Names

A data frame includes columns of fixed data types that are labeled with headers. You can identify the headers in a data frame using the `colnames` function. After reviewing the documentation for this function [here](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/row%2Bcolnames), please use the `colnames` function to extract the column names in the `df_participant` data frame.

```{r}
# TODO: Extract the column names of the `df_participant` table using the `colnames` function
col_names <- colnames(df_participant)

# Print the column names
print(col_names)
```

#### Identify Unique Participants

The built-in `unique` function is used to identify unique values in a data set. If the function is applied to a data frame, then unique rows will be returned. If the function is applied to a vector, then unique elements will be returned. Please review the documentation for the `unique` function [here](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/unique), then use the function to identify the unique values in the `id` column of `df_participant`.

```{r, max.height='200px'}
# TODO: Identify the unique values in the `id` column of `df_participant`
# Store the unique participant IDs in a vector
unique_participants <- unique(df_participant$id)

# Print the unique participants
print(unique_participants)
```

#### Count Unique Participants

The built-in `length` function can be used to identify the number of elements in a vector. Please review the documentation for the `length` function [here](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/length), then use the function to identify the number of unique participants in `df_participant`.

```{r}
# TODO: Use the `length` function to identify the number of unique participants in `df_participant`
n_participants <- length(unique_participants)

# Print the number of participants (There should be 9254 unique participants)
print(n_participants)
```

#### Sort Participant Identifiers

The built-in `sort` function can be used to sort values in a vector. When the values in a vector are alphanumeric, they are sorted in alphabetical order with numbers coming first. Please review the documentation for the `sort` function [here](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/sort), then use the function to sort the unique participant identifiers.

```{r, max.height='200px'}
# TODO: Sort the participants listed in `unique_participants` by their identifiers
unique_patients_sorted <- sort(unique_participants)

# Print the sorted participant identifiers
print(unique_patients_sorted)
```

#### Load Participant Encounters

Now, we will open a second table from our data set to inspect encounters with participants. Again using the `read.csv` function, please load the encounters data stored in `data/exams.csv`.

```{r, max.height='200px'}
# Specify the path to the encounters data
exams_path <- "data/exams.csv"

# TODO: Use the `read.csv`function to load the exam components into a data frame
df_exams <- read.csv(exams_path)
```

#### Define a Vector

We will use this notebook to inspect the number of exam components in each month. Please define a vector specifying the twelve months of the year ("January" through "December") as strings values.

```{r}
# TODO: Generate a vector of strings specifying the twelve months of the year
months <- c(
  "January", "February", "March", "April", "May", "June", "July", "August",
  "September", "October", "November", "December"
)
```

#### Extract Encounter Dates

Using the code provided below, we will extract the date of each encounter. We will get this information from the `exam_start` column of the `encounters` table. We will format the encounter dates as POSIXct (date and time) information. POSIXct internally represents date and time as seconds since Junuary 1, 1970. We use the `as.POSIXct` function (with documentation [here](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/as.POSIX*)) to convert string timestamps into POSIXct format. You have no task in the subsequent code block.

```{r, max.height='200px'}
# Extract encounters for each participant
encounters <- df_exams[, c("participant_id", "exam_start")]

# Reformat `exam_start` column as `POSIXct` (datetime) information
encounters$exam_start <- as.POSIXct(encounters$exam_start)

# Print the encounter start times
print(encounters)
```

#### Identify the Month of Each Encounter

For each of the exam components, we will identify the month of that encounter. We will do so by reformatting the data in the `exam_start` column using the `format` function (with documentation found [here](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/format)). The "%B" identifier specifies unabbreviated months (as noted [here](https://www.statmethods.net/input/dates.html). We will store the month in a new column. You have no task in the subsequent code block.

```{r, max.height='200px'}
# Extract the month for each encounter
encounters$MONTH <- format(encounters$exam_start, "%B")

# Display the resulting table
print(encounters)
```

#### Count Month Frequencies

We will loop through each month in the `months` vector. For each month, we will do the following:

1. Filter the rows in `encounters` to those with a corresponding value in the `MONTH` column
2. Count the number of rows in the filtered table
3. Store the row count in a dictionary called `n_encounters_by_month`

This will give us the number of encounters in each month. You have no task in the subsequent code block.

```{r}
# Initialize a list for encounter counts associated with each month
n_encounters_by_month <- list()

# Loop over months
for (month in months) {
  
  # Filter rows in `encounters` matching month
  encounters_filtered <- encounters[encounters$MONTH == month, ]
  
  # Count the number of rows in `encounters_filtered`
  nrows <- nrow(encounters_filtered)
  
  # Store the row count in `n_encounters_by_month`
  n_encounters_by_month[[month]] <- nrows
  
  # Print the number of encounters associated with the month
  cat("Month:", month, "\n")
  cat("Number of encounters:", nrows, "\n\n")
}
```

#### Summary

In this notebook, we loaded participant records into a data frame. We identified and sorted the unique participant IDs in our records. We also identified and plotted the number of exam components by month.
