---
title: Intro and Plotting
author: Alejandro Schuler
format:
  revealjs:
    smaller: true
    scrollable: false
    html-math-method: mathjax
error: true
echo: true
---

## Learning Goals:

- issue commands to R using the Rstudio REPL interface
- load a package into R
- read some tabluar data into R
- visualize tabluar data using ggplot geoms, aesthetics, and facets

# Basics {data-background-color=black}

## Console
The R console window is the left (or lower-left) window in RStudio.
The R console uses a "read, eval, print" loop. This is sometimes
called a REPL.

- Read: R reads what you type ...
- Eval: R evaluates it ...
- Print: R prints the result ...
- Loop: (repeat forever)

##

- The box contains an expression that will be evaluated by R, followed by the result of evaluating that expression.
```{r}
1 + 2
```
- `3` is the answer
- Ignore the `[1]` for now. 

- R performs operations (called *functions*) on data and values
- These can be composed arbitrarily
```{r}
log(1+3)
paste("The answer is", log(1+3))
```

## How do I...

- typing `?function_name` 
gives you information about what the function does
- Google is your friend. Try "function_name R language" or "how do I X in R?". I also strongly recommend using "tidyverse" in your queries or the name of a tidyverse package (more in a moment) that has related functions
- stackoverflow is your friend. It might take some scrolling, but you will eventually find what you need
- LLMs...

## Why learn this when LLMs can do it?

- Anthropic's Claude code (using eg Opus 5) or OpenAI's codex (using eg Sol) can do everything you will learn about in this class and probably also in any quant class you'll take in your MPH.

## Why learn this when LLMs can do it? {data-background-color=teal}

Discuss in your groups.

## Quadratic Equation {data-background-color=teal}

Solutions to a polynomial equation $ax^2 + bx + c = 0$ are given by
$$x = \frac{-b \pm \sqrt{b^2 - 4ac}}{2a}$$

Figure out how to use R functions and operations for square roots, exponentiation, and multiplication to calculate x given a=3, b=14, c=-5 (no LLMs).

```{r, echo=F, eval=F}
(-14 + sqrt(14^2 - 4*3*(-5)))/(2*3)
```

. . . 

- How did this **feel**? What was your emotional reaction when you saw the question?
- What did you **learn**? What did you notice?

## Packages
- The amazing thing about programming is that you are not limited to what is built into the language
- Millions of R users have written their own functions that you can use
- These are bundled together into *packages*
- To use functions that aren't built into the "base" language, you have to tell R to first go download the relevant code, and then to load it in the current session


```{r, eval=F}
install.packages("QuadRoot")
```
```{r}
library(QuadRoot)
QuadRoot(c(3,14,-5))
```

## 
- The `tidyverse` package has a function called `read_csv()` that lets you read csv (comma-separated values) files into R. 
- csv is a common format for data to come in, and it's easy to export csv files from microsoft excel, for instance. 

```{r}
# I have a file called "lupusGenes.csv" on github that we can read from the URL 
genes = read_csv("https://tinyurl.com/4vjrbwce")
```

- This fails because I haven't yet installed and loaded the `tidyverse` package
```{r, eval=F}
install.packages("tidyverse") # go download the package called "tidyverse"- only have to do this once
library("tidyverse") # load the package into the current R session - do this every time you use R and need functions from this package
```
```{r, include=F}
library(tidyverse)
```

```{r, eval=F}
genes = read_csv("https://tinyurl.com/4vjrbwce")
```

- Now there is no error message

## 
- packages only need to be loaded once per R session (session starts when you open R studio, ends when you shut it down)
- once the package is loaded it doesn't need to be loaded again before each function call
```{r, eval=F}
poly = read_csv("https://tinyurl.com/y22p5j6j") # reading another csv file
```


# Visualizing Data {data-background-color=black}

## Data analysis workflow
1. Read data into R (done!)
2. ~~Manipulate data~~
3. Get results, **make plots and figures**

## Getting your data in R
- Getting your data into R is easy. We already saw, for example:

```{r echo = T, results = 'hide', message=FALSE}
genes = read_csv("https://tinyurl.com/4vjrbwce")
```

- `read_csv()` requires you to tell it where to find the file you want to read in
  - Windows, e.g.: `"C:\Users\me\Desktop\myfile.csv"`
  - Mac, e.g.: `"/Users/me/Desktop/myfile.csv"`
  - Internet, e.g.: `"http://www.mywebsite.com/myfile.csv"`
- If your data is not already in csv format, google "covert X format to csv" or "read X format data in R"
- We'll learn the details of this later, but this is enough to get you started! 

## Looking at data
- `genes` is now a dataset loaded into R. To look at it, just type
```{r}
genes
```

This is a **data frame**, one of the most powerful features in R (a "tibble" is a kind of data frame).
- Similar to an Excel spreadsheet.
- One row ~ one
instance of some (real-world) object.
- One column ~ one variable, containing the values for the
corresponding instances.
- All the values in one column should be of the same type (a number, a category, text, etc.), but
different columns can be of different types.

## The Dataset
```{r}
genes
```

This is a subset of a real RNA-seq (GSE112087) dataset comparing RNA levels in blood between lupus (SLE) patients and healthy controls.

- 29 SLE Patients, 30 Healthy Controls
- We have basic metadata as well as the levels of multiple genes in blood.
- Let's see if we can find anything interesting from this already-generated data!

## Investigating a relationship

Let's say we're curious about the relationship between two genes RSAD2 and IFI44.

- Can we use R to make a plot of these two variables?

:::: {.columns}

::: {.column width="60%"}
```{r, tidy=F, out.height="80%"}
ggplot(genes) + 
  geom_point(aes(x = RSAD2, y = IFI44))
```
:::

::: {.column width="40%"}
- `ggplot(dataset)` says "start a chart with this dataset"
- `+ geom_point(...)` says "put points on this chart"
- `aes(x=x_values y=y_values)` says "map the values in the column `x_values` to the x-axis, and map the values in the column `y_values` to the y-axis" (`aes` is short for *aesthetic*)
:::

::::


## ggplot
:::: {.columns}

::: {.column width="60%"}
```{r, tidy=F}
ggplot(genes) + 
  geom_point(aes(x = VAPA, y = EIF3L))
```
:::

::: {.column width="40%"}
- `ggplot` is short for "grammar of graphics plot"
  - This is a language for describing how data get linked to visual elements
- `ggplot()` and `geom_point()` are functions imported from the `ggplot2` package, which is one of the "sub-packages" of the `tidyverse` package we loaded earlier
:::

::::

## Exercise: comparing expression levels {data-background-color=teal}

Make a scatterplot of `phenotype` vs `IFI44` (another gene in the dataset). The result should look like this:

```{r, tidy=F, echo=F}
ggplot(genes) + 
  geom_point(aes(x = phenotype, y = IFI44))
```

## Investigating a relationship
Let's say we're curious about the relationship between RSAD2 and IFI44.

```{r, tidy=F, echo=F}
ggplot(genes) + 
  geom_point(aes(x = RSAD2, y = IFI44, color = phenotype)) +
  theme(legend.position = "none")
```

- What's going on here? It seems like there are two clusters. 
- What is driving this clustering? Age? Sex? Ancestry? Phenotype?

## Aesthetics
- Aesthetics aren't just for mapping columns to the x- and y-axis
- You can also use them to assign color, for instance

```{r, tidy=F, eval=F}
ggplot(genes) + 
  geom_point(aes(x = RSAD2, 
                 y = IFI44,
                 color = phenotype))
```

```{r, tidy=F, echo=F}
ggplot(genes) + 
  geom_point(aes(x = RSAD2, 
                 y = IFI44,
                 color = phenotype))
```

***
- Aesthetics aren't just for mapping columns to the x- and y-axis
- We could have used a shape

```{r, tidy=F}
ggplot(genes) + 
  geom_point(aes(
    x = RSAD2, 
    y = IFI44, 
    shape=phenotype
  )) 
        
```

***
- Or size

```{r, tidy=F}
ggplot(genes) + 
  geom_point(aes(
    x = RSAD2, 
    y = IFI44, 
    size=ancestry
  )) 
```
- This one doesn't really make sense because we're mapping a categorical variable to an aesthetic that can take continuous values that imply some ordering


***
- If we set a property *outside* of the aesthetic, it no longer maps that property to a column. 

```{r, tidy=F}
ggplot(genes) + 
  geom_point(
    aes(
      x = RSAD2, 
      y = IFI44
    ),
    color = "blue"
  ) 
```
- However, we can use this to assign fixed properties to the plot that don't depend on the data


## Exercise: Plot {data-background-color=teal}

Can you recreate this plot?

```{r, tidy=F, echo=F, fig.width=16, fig.height=10, fig.align="center"}
ggplot(genes) + 
  geom_point(
    aes(
      x = RSAD2, 
      y = IFI44,
      color = phenotype,
      size = EIF3L
    )
  ) 
# increase legend and point size
```

## Exercise [together] {data-background-color=teal}

What will this do? Why?

```{r, tidy=F, eval=F}
ggplot(genes) + 
  geom_point(aes(x = RSAD2, y = IFI44, color = "blue"))
```

## Geoms

:::: {.columns}
::: {.column}
```{r, tidy=F}
ggplot(genes) + 
  geom_point(aes(x = RSAD2, y = IFI44))
```
:::
:::  {.column}
```{r, tidy=F, message=F}
ggplot(genes) + 
  geom_smooth(aes(x = RSAD2, y = IFI44))
```
:::
::::

- Both these plots represent the same data, but they use a different geometric representation ("geom")
- e.g. bar chart vs. line chart, etc. 
- R graph gallery is a great resource to help you design your plot and pick the right geom: r-graph-gallery.com


***
- Different geoms are configured to work with different aesthetics. 
- e.g. you can set the shape of a point, but you can’t set the “shape” of a line.
- On the other hand, you *can* set the "line type" of a line:

```{r, tidy=F, message=F}
ggplot(genes) + 
  geom_smooth(aes(x = RSAD2, y = IFI44, linetype = phenotype))
```

- Use e.g. `?geom_point` to see what aesthetics are expected or allowed

***
- It's possible to add multiple geoms to the same plot
```{r, tidy=F, message=F}
ggplot(genes) + 
  geom_smooth(aes(x = RSAD2, y = IFI44, color = phenotype)) + 
  geom_point(aes(x = RSAD2, y = IFI44, color = phenotype))
```

***
- To assign the same aesthetics to all geoms, pass the aesthetics to the `ggplot` function directly instead of to each geom individually
```{r, tidy=F, message=F}
ggplot(genes, aes(x = RSAD2, y = IFI44, color = phenotype)) + 
  geom_smooth() + 
  geom_point()
```

***
- You can also use different mappings in different geoms
```{r, tidy=F, message=F}
ggplot(genes, mapping = aes(x = RSAD2, y = IFI44)) + 
  geom_point(aes(color = ancestry)) + 
  geom_smooth()
```


## Exercise {data-background-color=teal}

Use google or other resources to figure out how to receate this plot in R:

```
ggplot(genes) + 
  ...
```

```{r, echo=F, tidy=F, message=F}
ggplot(genes) + 
  geom_bar(aes(x = ancestry, fill = phenotype))
```

- What might the name of this geom be? What properties of the plot (aesthetics) are mapped to what columns of the data?
- If you accomplish making the plot, can you figure out how to change the colors of the groups?

## Facets

- Aesthetics are useful for mapping columns to particular properties of a single plot
- Use **facets** to generate multiple plots with shared structure

```{r, tidy=F}
ggplot(genes) + 
  geom_point(aes(x = RSAD2, y = IFI44)) + 
  facet_wrap(vars(phenotype), nrow = 2)
```
- `facet_wrap` is good for faceting according to unordered categories

***
- `facet_grid` is better for ordered categories, and can be used with two variables

```{r, tidy=F}
ggplot(genes) + 
  geom_point(aes(x = RSAD2, y = IFI44)) + 
  facet_grid(rows=vars(phenotype), cols=vars(gender))
```


## Exercise {data-background-color=teal}

Use ggplot to investigate the relationship between gene expression and lupus using any combination of any kinds of plots that you like. Which genes are most associated with lupus? Does this vary by ancestry, age, or assigned sex at birth?

For some plots it may be helpful to reformat your data using this code (we'll learn how to do this on day 4):
```{r}

gene_names = names(genes)[6:11]
reformatted_genes = pivot_longer(genes, all_of(gene_names), names_to='gene', values_to='expression')
reformatted_genes
```
